Hochschild cohomology of torus-equivariant D-modules

### 2017-05-30 paper.tex
p₁ <-> p₂

[$p₂$ <-> $p₁$]
[subvariety <-> subscheme]
[macros]
[-advisor]
[map in contraction principle]

@@

By Verdier duality and adjunction we can always rewrite \eqref{eq:intro:d-mod-hcoh} as
\[
    \HCoh(\catDMod{\stack X}) \cong
    \opalg{\Hom_{\catDMod{X}}(k_{\stack X}, Δ^!Δ_!k_{\stack X})} =
    \opalg{\GammadR(\stack X, Δ^!Δ_!k_{\stack X})}.
\]
It is now tempting to look at the Cartesian square
\[
    \begin{tikzcd}
        \ls \stack X \arrow[r, "p₁"] \arrow[d, "p₂"] & \stack X \arrow[d, "Δ"] \\
        \stack X \arrow[r, "Δ"] & \stack X × \stack X
    \end{tikzcd}
\]
where
\[
    \ls \stack X = \stack X \mathop{×}\limits_{\stack X × \stack X} \stack X
\]
is the derived loop space (or inertia stack) of $\stack X$ and try to express the Hochschild cohomology as the cohomology of some sheaf on $\ls \stack X$.
We could expect the existence of an isomorphism
\begin{equation}
    \label{eq:intro:naive}
    \GammadR(\stack X, Δ^!Δ_!k_{\stack X}) \cong
    \GammadR(\stack X, p_{2,!}p₁^! k_{\stack X}).
\end{equation}
We refer to this as the \enquote{naive expectation}.
Unfortunately, the two sides are in general not isomorphic (the stack $\stack X = \ps1/\Ga$ is an easy counter-example).

In this paper we investigate how to quantify the cone of the morphism
\begin{equation}\label{eq:intro:central_cone}
    p_{2,!}p₁^! k_{\stack X} → Δ^!Δ_! k_{\stack X}
\end{equation}
and thus the failure of the naive isomorphism \eqref{eq:intro:naive} to hold.
As an application, we prove the following theorem, giving a class of stacks where \eqref{eq:intro:naive} is indeed an isomorphism.

\begin{Thm}\label{thm:main}
    Let $G \cong \Gm^n$ be a torus acting locally linearly on a scheme $X$ of finite type over $k$.
    Then there is a canonical isomorphism of algebras
    \[
        \HCoh\bigl(\catDMod{X/G}\bigr)
        \cong
        \opalg{\GammadR\bigl(X/G,\,p_{2,!} p₁^! k_{X/G}\bigr)},
    \]
    where the algebra structure on $\GammadR\bigl(X/G,\,p_{2,!} p₁^! k_{X/G}\bigr)$ is induced by the groupoid structure on $\ls(X/G)$.
\end{Thm}

@@

Recall now that we assume $Δ$ to be schematic and that the dualizing module $ω_{\stack X}$ is always holonomic.
Thus we have $\dualize Δ_* ω_{\stack X} = Δ_! k_{\stack X}$.
With this we observe that
\begin{align*}
    \HCoh\bigl(\catDMod{\stack X}\bigr)
    & = \Hom_{\catDMod{\stack X × \stack X}}(Δ_* ω_{\stack X},\, Δ_* ω_{\stack X}) & &\text{(Lemma~\ref{lem:pre:hcoh})} \\
    & = \opalg{\Hom_{\catDMod{\stack X × \stack X}}(Δ_! k_{\stack X},\, Δ_! k_{\stack X})} & & \text{(duality)} \\
    & = \opalg{\Hom_{\catDMod{\stack X × \stack X}}(k_{\stack X},\, Δ^! Δ_! k_{\stack X})} & & \text{(adjunction)} \\
    & = \opalg{\GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr)},
\end{align*}
where the algebra structure on $\GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr) = \Hom_{\catDMod{\stack X × \stack X}}(k_{\stack X},\, Δ^! Δ_! k_{\stack X})$ is the one induced by the $(Δ_!,Δ^!)$-adjunction monad.
Consider the Cartesian square
\[
    \begin{tikzcd}
        \ls{\stack X} \arrow[r, "p₁"] \arrow[d, "p₂"] & \stack X \arrow[d, "Δ"] \\
        \stack X \arrow[r, "Δ"] & \stack X × \stack X
    \end{tikzcd}
\]
Let us assume for the moment that $Δ$ (and hence $p_i$) is proper.
Then $Δ_* = Δ_!$ and $p_{2,*} = p_{2,!}$ and by Section~\ref{sec:monads} we have an isomorphism of monads
\begin{equation}
    \label{eq:central_iso}
    p_{2,!} p₁^! → Δ^!Δ_!,
\end{equation}
which induces an isomorphism of algebras
\[
    \GammadR\bigl(\stack X,\, p_{2,!} p₁^! k_{\stack X}\bigr)
    →
    \GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr).
\]
Of course, if $X$ is not an algebraic space, then $Δ$ is not proper (nor is it in general smooth).
Thus in general \eqref{eq:central_iso} is not an isomorphism and there is no canonical structure of monad on $p_{2,!} p₁^!$.
We would like to apply Lemma~\ref{lem:pre:groupoid_monad_hol} to construct a monad in special cases.
Thus the goal of the next section is to give a criterion for the assumptions of Lemma~\ref{lem:pre:groupoid_monad_hol}, i.e.~for base change to hold.

\begin{Ex}
    The base change morphism \eqref{eq:central_iso} is also an isomorphism if $Δ$ is smooth.
    In particular this implies that the \enquote{naive expectation} holds for $\stack X = \B G$ for any algebraic group $G$, i.e.~we have
    \[
        \HCoh(\catDMod{\B G}) = \opalg{\GammadR\bigl(\B G,\, p_{2,!} p₁^! k_{\B G}\bigr)}.
    \]
    An argument similar to \cite{BenZvi:mathoverflow:CohomologyOfGmodG} shows that there is a further isomorphism
    \[
        \GammadR\bigl(\B G,\, p_{2,!} p₁^! k_{\stack X}\bigr) \cong
        \GammadR(G, k_G)^\dual \otimes \GammadR(\B G, k_{\B G}).
    \]
    Alternatively, we can use the identification
    \[
        \catDMod{\B G} \cong \catModules{\GammadR(G, k_{G})^\dual},
    \]
    where the algebra structure on $\GammadR(G, k_G)^\dual$ is induced by the group multiplication \cite[Section~7.2]{DrinfeldGaitsgory:2013:FinitenessQuestions}.
    If $G$ is reductive, then $\GammadR(G, k_G)^\dual$ is an exterior algebra and thus its Hochschild cohomology can be computed directly.
\end{Ex}

>>>>

[subvariety <-> subscheme]
[macros]
[-advisor]
[map in contraction principle]

@@

By Verdier duality and adjunction we can always rewrite \eqref{eq:intro:d-mod-hcoh} as
\[
    \HCoh(\catDMod{\stack X}) \cong
    \opalg{\Hom_{\catDMod{X}}(k_{\stack X}, Δ^!Δ_!k_{\stack X})} =
    \opalg{\GammadR(\stack X, Δ^!Δ_!k_{\stack X})}.
\]
It is now tempting to look at the Cartesian square
\[
    \begin{tikzcd}
        \ls \stack X \arrow[r, "p₂"] \arrow[d, "p₁"] & \stack X \arrow[d, "Δ"] \\
        \stack X \arrow[r, "Δ"] & \stack X × \stack X
    \end{tikzcd}
\]
where
\[
    \ls \stack X = \stack X \mathop{×}\limits_{\stack X × \stack X} \stack X
\]
is the derived loop space (or inertia stack) of $\stack X$ and try to express the Hochschild cohomology as the cohomology of some sheaf on $\ls \stack X$.
We could expect the existence of an isomorphism
\begin{equation}
    \label{eq:intro:naive}
    \GammadR(\stack X, Δ^!Δ_!k_{\stack X}) \cong
    \GammadR(\stack X, p_{1,!}p₂^! k_{\stack X}).
\end{equation}
We refer to this as the \enquote{naive expectation}.
Unfortunately, the two sides are in general not isomorphic (the stack $\stack X = \ps1/\Ga$ is an easy counter-example).

In this paper we investigate how to quantify the cone of the morphism
\begin{equation}\label{eq:intro:central_cone}
    p_{1,!}p₂^! k_{\stack X} → Δ^!Δ_! k_{\stack X}
\end{equation}
and thus the failure of the naive isomorphism \eqref{eq:intro:naive} to hold.
As an application, we prove the following theorem, giving a class of stacks where \eqref{eq:intro:naive} is indeed an isomorphism.

\begin{Thm}\label{thm:main}
    Let $G \cong \Gm^n$ be a torus acting locally linearly on a scheme $X$ of finite type over $k$.
    Then there is a canonical isomorphism of algebras
    \[
        \HCoh\bigl(\catDMod{X/G}\bigr)
        \cong
        \opalg{\GammadR\bigl(X/G,\,p_{1,!} p₂^! k_{X/G}\bigr)},
    \]
    where the algebra structure on $\GammadR\bigl(X/G,\,p_{1,!} p₂^! k_{X/G}\bigr)$ is induced by the groupoid structure on $\ls(X/G)$.
\end{Thm}

@@

Recall now that we assume $Δ$ to be schematic and that the dualizing module $ω_{\stack X}$ is always holonomic.
Thus we have $\dualize Δ_* ω_{\stack X} = Δ_! k_{\stack X}$.
With this we observe that
\begin{align*}
    \HCoh\bigl(\catDMod{\stack X}\bigr)
    & = \Hom_{\catDMod{\stack X × \stack X}}(Δ_* ω_{\stack X},\, Δ_* ω_{\stack X}) & &\text{(Lemma~\ref{lem:pre:hcoh})} \\
    & = \opalg{\Hom_{\catDMod{\stack X × \stack X}}(Δ_! k_{\stack X},\, Δ_! k_{\stack X})} & & \text{(duality)} \\
    & = \opalg{\Hom_{\catDMod{\stack X × \stack X}}(k_{\stack X},\, Δ^! Δ_! k_{\stack X})} & & \text{(adjunction)} \\
    & = \opalg{\GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr)},
\end{align*}
where the algebra structure on $\GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr) = \Hom_{\catDMod{\stack X × \stack X}}(k_{\stack X},\, Δ^! Δ_! k_{\stack X})$ is the one induced by the $(Δ_!,Δ^!)$-adjunction monad.
Consider the Cartesian square
\[
    \begin{tikzcd}
        \ls{\stack X} \arrow[r, "p₂"] \arrow[d, "p₁"] & \stack X \arrow[d, "Δ"] \\
        \stack X \arrow[r, "Δ"] & \stack X × \stack X
    \end{tikzcd}
\]
Let us assume for the moment that $Δ$ (and hence $p_i$) is proper.
Then $Δ_* = Δ_!$ and $p_{1,*} = p_{1,!}$ and by Section~\ref{sec:monads} we have an isomorphism of monads
\begin{equation}
    \label{eq:central_iso}
    p_{1,!} p₂^! → Δ^!Δ_!,
\end{equation}
which induces an isomorphism of algebras
\[
    \GammadR\bigl(\stack X,\, p_{1,!} p₂^! k_{\stack X}\bigr)
    →
    \GammadR\bigl(\stack X,\, Δ^! Δ_! k_{\stack X}\bigr).
\]
Of course, if $X$ is not an algebraic space, then $Δ$ is not proper (nor is it in general smooth).
Thus in general \eqref{eq:central_iso} is not an isomorphism and there is no canonical structure of monad on $p_{1,!} p₂^!$.
We would like to apply Lemma~\ref{lem:pre:groupoid_monad_hol} to construct a monad in special cases.
Thus the goal of the next section is to give a criterion for the assumptions of Lemma~\ref{lem:pre:groupoid_monad_hol}, i.e.~for base change to hold.

\begin{Ex}
    The base change morphism \eqref{eq:central_iso} is also an isomorphism if $Δ$ is smooth.
    In particular this implies that the \enquote{naive expectation} holds for $\stack X = \B G$ for any algebraic group $G$, i.e.~we have
    \[
        \HCoh(\catDMod{\B G}) = \opalg{\GammadR\bigl(\B G,\, p_{1,!} p₂^! k_{\B G}\bigr)}.
    \]
    An argument similar to \cite{BenZvi:mathoverflow:CohomologyOfGmodG} shows that there is a further isomorphism
    \[
        \GammadR\bigl(\B G,\, p_{1,!} p₂^! k_{\stack X}\bigr) \cong
        \GammadR(G, k_G)^\dual \otimes \GammadR(\B G, k_{\B G}).
    \]
    Alternatively, we can use the identification
    \[
        \catDMod{\B G} \cong \catModules{\GammadR(G, k_{G})^\dual},
    \]
    where the algebra structure on $\GammadR(G, k_G)^\dual$ is induced by the group multiplication \cite[Section~7.2]{DrinfeldGaitsgory:2013:FinitenessQuestions}.
    If $G$ is reductive, then $\GammadR(G, k_G)^\dual$ is an exterior algebra and thus its Hochschild cohomology can be computed directly.
\end{Ex}
